Reactive SINDy: Discovering governing reactions from concentration data

### 2018-12-08 docs/manuscript_jcp.tex
wip

Here we model the MAPK pathway with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network consists of activation/phosphorylation reactions
\begin{align*}
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    } \\
\end{align*}
and deactivation/dephosphorylation reactions
\begin{align*}
\mathrm{MAPKKK*}                     & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}                      & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}                       & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}                         & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the response, i.e. the concentration $[\mathrm{TF*}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.

We generate concentration time series data of the MAPK reactions above at 
three different initial conditions, each differing in the amount of stimulus $[\mathrm{S}]$.
They are marked in Fig.~\ref{fig:mapk}a by vertical dashed lines.
These time series are concatenated to yield a dataset of 300 frames in total.
We use the library $\Theta$ of candidate reactions Tab.~\ref{tab:reaction-library-mapk}.
To determine the hyperparameter $\alpha=6.6\times 10^{-9}$ we shuffle the data and perform
15-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:mapk}b.
Least-squares estimation detects 5 of the 8 reaction processes that belong to the 
ground truth model. However it also detects 12 other reaction 
processes ($\theta_{18}$ - $\theta_{29}$) which are not in the ground truth.
Reactive SINDy estimation detects all reactions of the ground truth, 
two processes ($\theta_{4}$ and $\theta_{8}$) show deviations in 
rate constants. Generally reactive SINDy yields a sparse model which allows further
simplification of the reaction network by dropping out reaction processes that lie 
beneath a certain cutoff. In this case for example a cutoff of $0.25$ would directly recover
the ground truth reaction network.
Quantitatively, one may consider the $L_1$ norm of the relative distance 
of estimated rate constants $\hat{\xi}_r$ to the non-zero rate 
constants of the ground truth $\xi_r$
\begin{equation*}
\sum_{r=1}^{8} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $167\%$ error for least-squares and $21\%$ error for reactive SINDy.

@@

\subsection{Application to Lotka-Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka-Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reactions
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generate concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. We construct the library $\Theta$ 
of candidate reactions given in Tab.~\ref{tab:reaction-library-lv}. 
We determine the hyperparameter $\alpha=2.7\times10^{-7}$ by shuffling 
the data and performing 5-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:lv}b.
Least-squares estimation detects all reactions of the ground truth model but also
two false processes ($\theta_{6}$ and $\theta_{7}$) with a higher rate than
the two first true processes ($\theta_{1}$ and $\theta_{2}$). Reactive SINDy detects
the true reaction network with minor deviations in rate constants. As in 
Sec.~\ref{sec:mapk} we consider the $L_1$ norm of the relative distance to the ground truth
for non-zero rate constants
\begin{equation*}
\sum_{r=1}^{5} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $75\%$ error for least-squares and $7\%$ error for reactive SINDy.

>>>>

Here the MAPK pathway is modeled with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network consists of activation/phosphorylation reactions
\begin{align*}
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    } \\
\end{align*}
and deactivation/dephosphorylation reactions
\begin{align*}
\mathrm{MAPKKK*}                     & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}                      & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}                       & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}                         & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the response, i.e., the concentration $[\mathrm{TF*}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.

We generate concentration time series data of the MAPK reactions above at 
three different initial conditions, each differing in the amount of stimulus $[\mathrm{S}]$.
The response yielded by the chosen initial conditions is marked in Fig.~\ref{fig:mapk}a by vertical dashed lines.
The concatenated time series is a dataset of 300 frames in total.
We use the library $\Theta$ of ansatz reactions Tab.~\ref{tab:reaction-library-mapk}.
The hyperparameter $\alpha=6.6\times 10^{-9}$ was determined by shuffling the data and performing
15-fold cross validation. The estimated 
rate constants were obtained by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:mapk}b.
Least-squares estimation detects 5 of the 8 reaction processes that belong to the 
ground truth model. However it also detects 12 spurious reaction 
processes ($\theta_{18}$ - $\theta_{29}$).
Reactive SINDy estimation detects all reactions of the ground truth, 
two processes ($\theta_{4}$ and $\theta_{8}$) show deviations in 
rate constants. Generally reactive SINDy yields a sparse model which allows further
simplification of the reaction network by dropping out reaction processes that lie 
beneath a certain cutoff. In this case for example a cutoff of $\kappa=0.25$ would directly recover
the ground truth reaction network.
Quantitatively, one may consider the $L_1$ norm of the relative distance 
of estimated rate constants $\hat{\xi}_r$ to the non-zero rate 
constants of the ground truth $\xi_r$
\begin{equation*}
\sum_{r=1}^{8} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $167\%$ error for least-squares and $21\%$ error for reactive SINDy.

@@

\subsection{Application to Lotka--Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka--Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reaction network
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generate concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. We construct the library $\Theta$ 
of candidate reactions given in Tab.~\ref{tab:reaction-library-lv}. 
We determine the hyperparameter $\alpha=2.7\times10^{-7}$ by shuffling 
the data and performing 5-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:lv}b.
Least-squares estimation detects all reactions of the ground truth model but also
two false processes ($\theta_{6}$ and $\theta_{7}$) with a higher rate than
the two first true processes ($\theta_{1}$ and $\theta_{2}$). Reactive SINDy detects
the true reaction network with minor deviations in rate constants. As in 
Sec.~\ref{sec:mapk} we consider the $L_1$ norm of the relative distance to the ground truth
for non-zero rate constants
\begin{equation*}
\sum_{r=1}^{5} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $75\%$ error for least-squares and $7\%$ error for reactive SINDy.
